CS267 Final Project: An Empirical Investigation of the Algorithmic Lov\'{a}sz Local Lemma

### 2015-05-19 paper/cs267-report/paper.tex
Add another paper and a poster.

\documentclass[twocolumn]{article}
\usepackage{amsmath, amssymb}
\usepackage[retainorgcmds]{IEEEtrantools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{filecontents}
\usepackage{hyperref}
\usepackage{graphicx}
\author{Derek Kuo, Henry Milner}
\title{CS267 Final Project: An Empirical Investigation of the Algorithmic Lovász Local Lemma}
\date{5/11/15}

@@

\section{Introduction}
\label{sec:intro}
In 1975, Erd\"{o}s and \lovasz \cite{erdos1975problems} developed a technique, now called the \lovasz Local Lemma (LLL), for proving the existence of solutions to combinatorial problems.  The technique is a ``probabilistic method'': Given a set of events that we want to avoid (for example, each event could correspond to the violations of a problem constraint), it is proved that there is a strictly positive probability that none of these events happen under a random potential solution drawn from a suitable distribution on potential solutions.  The main precondition of the LLL is that the statistical dependencies among events are \emph{sparse}, with the sparsity being of the same magnitude as the marginal probability of each single event being unsatisfied under the sampling distribution.  We will discuss the technique later in more detail.

@@

To study applications of the algorithmic LLL, we must choose combinatorial problems to solve.  We focus on \ksat.  In \ksat, we are given $n$ boolean variables and $m$ disjunctive (``OR'') clauses of $k$ possibly-negated variables each, and we must find an assignment of the variables that satisfies all of the clauses.  The decision version of \ksat is NP-complete, but this does not necessarily mean that instances of practical importance (or, possibly, all but a small number of worst-case instances) are hard, and there is a long and ongoing history of work on heuristic search methods for the problem.  In fact, there is an annual competition for SAT solvers \cite{belov2014sat}.  (Our benchmarks will be based on the guidelines in that competition.)  We do not expect that our implementations will be competitive with state-of-the-art SAT solvers.  Our interest is primarily in finding out whether there is any practical potential for an algorithm whose development has thus far been guided exclusively by the desire to prove results about its performance.

@@

\section{The Algorithmic \lovasz Local Lemma}
\label{sec:alll}
LLL algorithms can be useful when a problem can be phrased as follows.  Let $\mathcal{A} = \{A_1, \cdots, A_n\} \in \{0,1\}^n$ be a set of discrete variables (which we take to be binary for sake of exposition), and let $\mathcal{E} = \{E_1, \cdots, E_m\}$ be a set of functions $\{0,1\}^n \to \{0,1\}$ mapping the variables to truth values.  The $E_i$ are called ``events,'' and $E_i$ is said to ``happen'' under some assignment to $\mathcal{A}$ if $E_i(\mathcal{A}) = 1$.  We would like to find an assignment of the variables so that none of the events happen.

This framework encompasses many constraint satisfaction problems.  For example, we can encode graph coloring as follows: the variables are binary encodings of the color assignments to vertices, and event $E_{ij}$ happens if vertices $i$ and $j$ are assigned the same color.  Then if we find an assignment of the variables for which none of the events happen, we have found a graph coloring.  Encoding \ksat is even more straightforward: the $A_i$ are the problem variables, and each $E_i$ corresponds to a clause involving $k$ variables.  $E_i$ happens if the corresponding clause is false, and an assignment solves the problem if none of the events happen.  In \ksat, each event $E_i$ is a function of only a smaller subset of the variables (of size $k$), and it is functionally independent of the rest.  We denote that subset of $\mathcal{A}$ by $\operatorname{vbl}(E_i)$.  If we imagine randomly and independently determining the variable values, then $E_i$ is statistically dependent only on $\operatorname{vbl}(E_i)$.  LLL algorithms typically need $|\operatorname{vbl}(E_i)| << |\mathcal{A}|$.

@@

A \ksat problem satisfies this condition if each clause shares variables with no more than $\frac{2^k}{e}$ other clauses.  For instance, if we generate a random \ksat instance with $n$ variables and $m$ clauses (and random signs), the probability that two clauses share a variable is approximately $\frac{k^2}{n}$, so the expected degree of any node is approximately $\frac{m k^2}{n}$.  If we make the optimistic assumption that the maximum degree equals the expected degree, then the proof by Moser and Tardos applies if $m$ is less than $\frac{2^k n}{k^2 e}$.  For comparison, the maximum number of clauses in a \ksat instance is $2^k {n \choose k}$.  It is known that random 3-SAT instances are hardest when $m \approxeq 4.26 n$ \cite{gomes2008satisfiability}, so these are relatively easy instances.  In fact, for $k = 3$ these instances do not even use all the variables, since $\frac{2^3 m}{3^2 e} = \frac{8}{9e} < 1/3$.  For larger $k$, the instances for which the proof by Moser and Tardos applies can be much less trivial than that, but they can never meet the threshold, which is asymptotically $\log 2 2^k - \frac{1}{2}(1 + \log 2)$\cite{coja2014asymptotic}.  Of course, we should not expect to be able provably to solve hard instances of NP-complete problems in polynomial time!  We will provide empirical results for problems that are at varying degrees of proximity to the hardness threshold.

>>>>

\documentclass[twocolumn]{article}
\usepackage{amsmath, amssymb}
\usepackage[retainorgcmds]{IEEEtrantools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{filecontents}
\usepackage{hyperref}
\usepackage{graphicx}
\author{Derek Kuo, Henry Milner}
\title{CS267 Final Project: An Empirical Investigation of the Algorithmic Lov\'{a}sz Local Lemma}
\date{5/11/15}

@@

\section{Introduction}
\label{sec:intro}
In 1975, Erd\"{o}s and \lovasz \cite{erdos1975problems} developed a technique, now called the \lovasz Local Lemma (LLL), for proving the existence of solutions to combinatorial problems.  The technique is a ``probabilistic method'': Given a set of events that we want to avoid (for example, each event could correspond to the violation of one problem constraint), it is proved that there is a strictly positive probability that none of these events if we draw randomly from a suitable distribution on potential solutions.  The main precondition of the LLL is that the statistical dependencies among events are \emph{sparse}, with the sparsity being of the same magnitude as the marginal probability of each single event being unsatisfied under the sampling distribution.  We will discuss the technique later in more detail.

@@

To study applications of the algorithmic LLL, we must choose a combinatorial problem to solve.  We focus on \ksat.  In \ksat, we are given $n$ boolean variables and $m$ disjunctive (``OR'') clauses of $k$ possibly-negated variables each, and we must find an assignment of the variables that satisfies all of the clauses.  The decision version of \ksat is NP-complete, but this does not necessarily mean that instances of practical importance (or, possibly, all but a small number of worst-case instances) are hard, and there is a long and ongoing history of work on heuristic search methods for the problem.  In fact, there is an annual competition for SAT solvers \cite{belov2014sat}.  (Our benchmarks will be based on the guidelines in that competition.)  We do not expect that our implementations will be competitive with state-of-the-art SAT solvers, and such comparisons are outside the scope of this project.  Our interest is primarily in finding out whether there is any practical potential for an algorithm whose development has thus far been guided exclusively by the desire to prove results about its performance.

@@

\section{The Algorithmic \lovasz Local Lemma}
\label{sec:alll}
LLL algorithms can be useful for problems that can be phrased as follows.  Let $\mathcal{A} = \{A_1, \cdots, A_n\} \in \{0,1\}^n$ be a set of discrete variables (which we take to be binary for sake of exposition), and let $\mathcal{E} = \{E_1, \cdots, E_m\}$ be a set of functions $\{0,1\}^n \to \{0,1\}$ mapping the variables to truth values.  The $E_i$ are called ``events,'' and $E_i$ is said to ``happen'' under some assignment to $\mathcal{A}$ if $E_i(\mathcal{A}) = 1$.  We would like to find an assignment of the variables so that none of the events happen.

This framework encompasses many constraint satisfaction problems.  For example, we can encode graph coloring as follows: the variables are binary encodings of the color assignments to vertices, and event $E_{ij}$ happens if vertices $i$ and $j$ are assigned the same color.  Then if we find an assignment of the variables for which none of the events happen, we have found a graph coloring.  Encoding \ksat is even more straightforward: the $A_i$ are the problem variables, and each $E_i$ corresponds to a clause involving $k$ variables.  $E_i$ happens if the corresponding clause is false, and an assignment solves the problem if none of the events happen.  In \ksat, each event $E_i$ is a function of only a smaller subset of the variables (of size $k$), and it is functionally independent of the rest.  We denote that subset of $\mathcal{A}$ by $\operatorname{vbl}(E_i)$.  If we imagine randomly and independently determining the variable values, then $E_i$ is statistically dependent only on $\operatorname{vbl}(E_i)$.  LLL algorithms typically need $|\operatorname{vbl}(E_i)| \ll |\mathcal{A}|$.

@@

A \ksat problem satisfies this condition if each clause shares variables with no more than $\frac{2^k}{e}$ other clauses.  For instance, if we generate a random \ksat instance with $n$ variables and $m$ clauses (and random signs), the probability that two clauses share a variable is approximately $\frac{k^2}{n}$, so the expected degree of any node is approximately $\frac{m k^2}{n}$.  If we make the optimistic assumption that the maximum degree equals the expected degree, then the proof by Moser and Tardos applies if $m$ is less than $\frac{2^k n}{k^2 e}$.  For comparison, the maximum number of clauses in a \ksat instance is $2^k {n \choose k}$.  It is known that random 3-SAT instances are hardest when $m \approxeq 4.26 n$ \cite{gomes2008satisfiability}, so these are relatively easy instances.  In fact, for $k = 3$ these instances do not even use all the variables, since $\frac{2^3 m}{3^2 e} = \frac{8}{9e} < 1/3$.  For larger $k$, the instances for which the proof by Moser and Tardos applies can be much less trivial than that, but they can never meet the threshold, which is asymptotically $(\log 2) 2^k - \frac{1}{2}(1 + \log 2)$\cite{coja2014asymptotic}.  Of course, we should not expect to be able provably to solve hard instances of NP-complete problems in polynomial time!  We will provide empirical results for problems that are at varying degrees of proximity to the hardness threshold.
